Avoiding deforestation: The implications for global markets

### 2024-02-28 paper.Rmd
Update paper.Rmd

---
title: "Avoiding deforestation: The implications for global markets"
date: "`r format(Sys.time(), '%B %d, %Y')`"
author: "Eric C. Davis, Maros Ivanic, and Brent Sohngen"
output: 
  bookdown::pdf_document2:
    keep_tex: TRUE
toc: FALSE
bibliography: references.bib
header-includes:
    - \usepackage{setspace}\doublespacing
abstract: "The projected growth in population and incomes as implied by various shared socioeconomic pathways is expected to create pressure to convert forestland into farmland. At the same time, the increasingly negative climate impacts are expected to generate increasing pressure to enhance the terrestrial carbon sink. These goals seem incompatible as reversing the deforestation trend by afforesting cropland might result in negative market impacts such as higher food prices, especially in low agricultural productivity/high population growth scenarios. In order to better grasp the interrelationship and the magnitude of harm that might be caused by prioritizing the goal of mitigating deforestation, this analysis connects two respected global economic models: the Global Trade Analysis Project (GTAP) model and the Global Timber Model (GTM). Results from the model suggest that it is possible through fiscal policy to preserve 144.2 million hectares of forestland that otherwise would be converted to agricultural land by 2033. As to the economic price for doing so, the avoided deforestation would in most regions of the world result in less agricultural output and higher market prices. This is estimated to impact the well-being of global consumers by $92.1 billion, which translates to an global average cost of $10.6 per person in 2033."
---

@@

The global gains that can be realized from afforestation, however, are uncertain, with broad estimates of cumulative sequestration ranging from 42 gigatonnes to 700 gigatonnes by the end of the century [@veldman2019; @bastin2019; @humpenoder2014; @sathaye2006; @doelman2020]. In addition, the amount of land conversion that would be required also varies greatly, with estimates ranging from 0.3 billion hectares to 2.8 billion hectares [@veldman2019; @bastin2019; @humpenoder2014; @doelman2020]. In many cases, these results suggest massive losses in the amount of pasture and cropland that would be available. This puts into question the feasibility of attaining these results, as almost 90 percent of deforestation worldwide is undertaken to expand the amount of agricultural land [@fao2021]. There are many reasons for this. For example, as nations develop and disposable incomes rise, consumers often change to more meat-based diets. Population growth is also a driver. Currently, the United Nations' medium variant population projections suggest that between 2023 and 2033 the worlds' population will grow by 8.8 percent, a gain of 707.1 million people [@un2024]. And where population growth is especially high, which is the case at present throughout most of Africa (2023--2033 population growth: 25.0%: 360.6 million), this puts extreme pressure on policymakers to allow greater conversion of forestland into farmland. Still, some research has suggested that substantive gains in the carbon sink (+23.8PgCo2e/yr) can be made without sacrificing agricultural land [@griscom2017].

It is important to understand where to target afforestation efforts, as they are thought to not provide equal benefits in every region. For example, afforestation in boreal zones has been shown to provide the smallest net benefits, as the carbon sequestrations benefits of conifer forests are mitigated by the reduction in albedo in winter [@mykleby2017; @li2015]. Coniferous forests have also been shown to increase soil organic carbon less than broadleaf forests [@guo2021]. Conversely, tropical forests have been linked with a strong positive result from afforestation, as the carbon sequestration benefit is complemented by a cooling impact due to changes in both albedo and evapotranspiration, and with temperate forests, the impact of afforestation is also shown to be positive, albeit less strongly so due to a winter warming effect [@li2015].

@@

Toward this end, @steinbuks2012 developed their global partial equilibrium FABLE model (Forestry, Agriculture, Biofuels, Land Use, and Environment) to identify the optimal land use choices given the competing demands of meeting greenhouse gas targets and satisfying demand for food, bioenergy, forest products, and ecosystem services. While FABLE is a powerful tool for examining the optimal trajectory of various land uses under specific demand assumptions, it is focused on specific sectors of the economy and neglects general equilibrium and household welfare effects [@steinbuks2012; @hertel2017]. Other efforts have used the Global Trade Analysis Project (GTAP) computable general equilibrium model to address such questions, but its standard model does not well account for land-use changes. This prompted the creation of the Land Use and Land Cover Database within the GTAP framework, which incorporates forestry remote-sensing products [@baldos2012]. The GTAP Agro-Ecological Zone (GTAP-AEZ) model further strengthened the GTAP model's ability to analyze land use changes, with the use of spatially explicit global land use data and through the incorporation of intra- and inter-regional land and land-based greenhouse emissions heterogeneity [@hertel2008]. While the GTAP-AEZ model benefits from the inclusion of competition among different crops, grazing, and forest-based uses, it has limitations, primarily due to its infrequent updates. The KLUM@GTAP framework integrated the Kleines Land Use Model (KLUM) with an extended yet static version of GTAP called GTAP-EFL to evaluate the impact of climate change on cropland allocation. GTAP-EFL separates energy factors from intermediate inputs and incorporates them into capital, while also considering CO2 emissions. KLUM, a global agricultural land-use model, connects the economy to global crop allocation to maximize producer returns under specific risk assumptions. KLUM@GTAP substitutes the land allocation mechanism within GTAP-EFL by utilizing regionally aggregated area changes in cropland determined by KLUM to update cropland shares in GTAP-EFL [@ronneberger2009].

More recent modeling attempts of land-use change have begun to move away from comparative static analysis. For example, the dynamic GTAP model, supported by the GTAP-AEZ database, considers land market effects, which were identified as significant in driving results by @stevenson2013. This model has been utilized to evaluate the impact of crop intensification on land use. It, however, has been critiqued, as global aggregates may mask localized shifts that can have implications for ecologically significant areas [@byerlee2014]. In addition, in the standard GTAP model, it is difficult to model a sector individually. There are 68 market sectors covered in the latest version of the GTAP database. The GTAP model's general equilibrium nature means that all these markets are cleared with market-clearing prices. As the model applies nested CES production functions to all sectors and differentiates them by parameters only, to model one sector, in this case forestry, differently from the rest would require a major redesign of the model. While not impossible, this project will attempt to get around these concerns and assess the impact of various likely land-use scenarios by melding the dynamic GTAP model with the Global Timber Model (GTM), which is a dynamic optimization model that has been specifically designed to analyze the relationship among land rents, forested land cover, and carbon sequestration through forests and thus is better able to examine the impact of afforestation or avoided deforestation on the carbon sink [@sohngen2001]. The latest versions of both models were used. For the GTAP model, this meant using data which has 2017 as its base year. This data was then updated to the base year (2023) using data from the Centre for Prospective Studies and International Information (CEPII) [@CEPII2024] to account for the projected changes in gross domestic product (GDP), population, and productivity for each country in the world. Similar steps were taken to update and recalibrate the GTM model.

@@

To understand the impact of potential regional and global afforestation actions, an important first step was creating a realistic baseline scenario that depicts the likely changes over the decade under study, 2023--2033. Data for this were drawn from CEPII [@CEPII2024]. To build our scenario, we used two key projected variables by CEPII: population and GDP. The CEPII projections do not contain any information on land use. To build a more complete baseline, we also include the projections on forested land change included in the Global Timber Model (GTM). Using the GTAP model in connection with the GTM model (to assure that the changes in land are consistent between the two models), we calculate the overall change in factor quantity/productivity growth required to achieve the targeted values in 2033. In Table \@ref(tab:baseline-growth) we show the projected population and GDP growths per region. GDP increases for all regions of the world with the largest percentage changes occurring in South Asia (+308.6%), China (+189.8%), and Sub-Saharan Africa (+172.9%). Population also is projected to grow in most regions with the largest changes happening in Sub-Saharan Africa (+55.7%). Japan, Russia, and East Asia are projected to see their populations decline by 10.4%, 4.7%, and 2.6%, respectively. As for the projected change in agricultural land, baseline estimates show significant growth in Oceania (+27.6%), Rest of Latin America (+25.8%), Sub-Saharan Africa (+17.4%), Central America (+13.9%), and Brazil (+13.9%). Japan is the only nation that sees a significant decline (-34.9%) in agricultural land, which is likely tied in part to the population decline described above.

>>>>

---
title: "Avoiding deforestation: The implications for global markets"
date: "`r format(Sys.time(), '%B %d, %Y')`"
author: "Eric C. Davis, Maros Ivanic, and Brent Sohngen"
output: 
  bookdown::pdf_document2:
    keep_tex: TRUE
toc: FALSE
bibliography: references.bib
header-includes:
    - \usepackage{setspace}\doublespacing
abstract: "The projected growth in population and incomes as implied by various shared socioeconomic pathways is expected to create pressure to convert forestland into farmland. At the same time, the increasingly negative climate impacts are expected to generate further pressure to enhance the terrestrial carbon sink. These goals are potentially incompatible as reversing the deforestation trend by afforesting cropland might result in negative market impacts such as higher food prices, especially in low agricultural productivity/high population growth scenarios. In order to better grasp the interrelationship and the magnitude of harm that might be caused by prioritizing the goal of mitigating deforestation, this analysis connects two respected global economic models: the Global Trade Analysis Project (GTAP) model and the Global Timber Model (GTM). Results from the model suggest that it is possible through fiscal policy to preserve 144.2 million hectares of forestland that otherwise would be converted to agricultural land by 2033. As to the economic price for doing so, the avoided deforestation would in most regions of the world result in less agricultural output and higher market prices. This is estimated to impact the well-being of global consumers by $92.1 billion, which translates to a global average cost of $10.6 per person in 2033."
---

@@

The global gains that can be realized from afforestation, however, are uncertain, with broad estimates of cumulative sequestration ranging from 42 gigatonnes to 700 gigatonnes by the end of the century [@veldman2019; @bastin2019; @humpenoder2014; @sathaye2006; @doelman2020]. In addition, the amount of land conversion that would be required also varies greatly, with estimates ranging from 0.3 billion hectares to 2.8 billion hectares [@veldman2019; @bastin2019; @humpenoder2014; @doelman2020]. In many cases, these results suggest massive losses in the amount of pasture and cropland that would be available. This puts into question the feasibility of attaining these results, as almost 90 percent of deforestation worldwide is undertaken to expand the amount of agricultural land [@fao2021]. There are many reasons for this. For example, as nations develop and disposable incomes rise, consumers often change to more meat-based diets. Population growth is also a driver. Currently, the United Nations' medium variant population projections suggest that between 2023 and 2033 the worlds' population will grow by 8.8 percent, a gain of 707.1 million people [@un2024]. And where population growth is especially high, which is the case at present throughout most of Africa (2023 to 2033 population growth: 25.0%: 360.6 million), this puts extreme pressure on policymakers to allow greater conversion of forestland into farmland. Still, some research has suggested that substantive gains in the carbon sink (+23.8PgCo2e/yr) can be made without sacrificing agricultural land [@griscom2017].

If such efforts are to be made, it is important to understand where to target them, as afforestation is thought to not provide equal benefits in every region. For example, afforestation in boreal zones has been shown to provide the smallest net benefits, as the carbon sequestrations benefits of conifer forests are mitigated by the reduction in albedo in winter [@mykleby2017; @li2015]. Coniferous forests have also been shown to increase soil organic carbon less than broadleaf forests [@guo2021]. Conversely, tropical forests have been linked with a strong positive result from afforestation, as the carbon sequestration benefit is complemented by a cooling impact due to changes in both albedo and evapotranspiration, and with temperate forests, the impact of afforestation is also shown to be positive, albeit less strongly so due to a winter warming effect [@li2015].

@@

Several notable modeling attempts have been made in this direction. @steinbuks2012 developed their global partial equilibrium FABLE model (Forestry, Agriculture, Biofuels, Land Use, and Environment) to identify the optimal land use choices given the competing demands of meeting greenhouse gas targets and satisfying demand for food, bioenergy, forest products, and ecosystem services. While FABLE is a powerful tool for examining the optimal trajectory of various land uses under specific demand assumptions, it is focused on specific sectors of the economy and neglects general equilibrium and household welfare effects [@steinbuks2012; @hertel2017]. Other efforts have used the Global Trade Analysis Project (GTAP) computable general equilibrium model to address such questions, but its standard model does not well account for land-use changes. This prompted the creation of the Land Use and Land Cover Database within the GTAP framework, which incorporates forestry remote-sensing products [@baldos2012]. The GTAP Agro-Ecological Zone (GTAP-AEZ) model further strengthened the GTAP model's ability to analyze land use changes, with the use of spatially explicit global land use data and through the incorporation of intra- and inter-regional land and land-based greenhouse emissions heterogeneity [@hertel2008]. While the GTAP-AEZ model benefits from the inclusion of competition among different crops, grazing, and forest-based uses, it has limitations, primarily due to its infrequent updates. The KLUM@GTAP framework integrated the Kleines Land Use Model (KLUM) with an extended yet static version of GTAP called GTAP-EFL to evaluate the impact of climate change on cropland allocation. KLUM is a global agricultural land-use model that connects the economy to global crop allocation to maximize producer returns under specific risk assumptions, and GTAP-EFL separates energy factors from intermediate inputs and incorporates them into capital, while also considering CO2 emissions. KLUM@GTAP substitutes the land allocation mechanism within GTAP-EFL by utilizing regionally aggregated area changes in cropland determined by KLUM to update cropland shares in GTAP-EFL [@ronneberger2009].

More recent modeling attempts of land-use change have begun to move away from comparative static analysis though. For example, the dynamic GTAP model, supported by the GTAP-AEZ database, considers land market effects, which were identified as significant in driving results by @stevenson2013. This model has been utilized to evaluate the impact of crop intensification on land use. It, however, has been critiqued, as global aggregates may mask localized shifts that can have implications for ecologically significant areas [@byerlee2014]. In addition, in the standard GTAP model, it is difficult to model a sector individually. There are 68 market sectors covered in the latest version of the GTAP database. The GTAP model's general equilibrium nature means that all these markets are cleared with market-clearing prices. As the model applies nested CES production functions to all sectors and differentiates them by parameters only, to model one sector, in this case forestry, differently from the rest would require a major redesign of the model. While not impossible, this project will attempt to get around these concerns and assess the impact of various likely land-use scenarios by melding the dynamic GTAP model with the Global Timber Model (GTM), which is a dynamic optimization model that has been specifically designed to analyze the relationship among land rents, forested land cover, and carbon sequestration through forests and thus is better able to examine the impact of afforestation or avoided deforestation on the carbon sink [@sohngen2001]. The latest versions of both models were used. For the GTAP model, this meant using data which has 2017 as its base year. This data was then updated to the base year (2023) using data from the Centre for Prospective Studies and International Information (CEPII) [@CEPII2024] to account for the projected changes in gross domestic product (GDP), population, and productivity for each country in the world. Similar steps were taken to update and recalibrate the GTM model.

@@

To understand the impact of potential regional and global afforestation actions, an important first step was creating a realistic baseline scenario that depicts the likely changes over the decade under study, 2023--2033. Data for this were drawn from CEPII [@CEPII2024]. To build our scenario, we use two of their variables that we determined to be key: population and GDP. The CEPII projections do not contain any information on land use. To build a more complete baseline, we also include the projections on forested land change included in the Global Timber Model (GTM). Using the GTAP model in connection with the GTM model (to assure that the changes in land are consistent between the two models), we calculate the overall change in factor quantity/productivity growth required to achieve the targeted values in 2033. In Table \@ref(tab:baseline-growth), we show the projected population and GDP growth per region. GDP increases for all regions of the world with the largest percentage changes occurring in South Asia (+308.6%), China (+189.8%), and Sub-Saharan Africa (+172.9%). Population also is projected to grow in most regions with the largest changes happening in Sub-Saharan Africa (+55.7%). Japan, Russia, and East Asia are projected to see their populations decline by 10.4%, 4.7%, and 2.6%, respectively. As for the projected change in agricultural land, our baseline prjoections show significant growth in Oceania (+27.6%), Rest of Latin America (+25.8%), Sub-Saharan Africa (+17.4%), Central America (+13.9%), and Brazil (+13.9%). Japan is the only nation that sees a significant decline (-34.9%) in agricultural land, which is likely tied in part to the population decline just described.
